MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-04-02 MAESTROeX_I/algorithm.tex

\subsection{Spatial Discretization}\label{Sec:Spatial}
MAESTRO contains variables defined on Cartesian grids 
one-dimensional base state stored on radial arrays.

@@

\subsection{Temporal Integration Scheme}\label{Sec:Temporal Integration Scheme}
Previously we adopted an approach where we split the velocity into a base state component, $w_0(r,t)$, 
and a local velocity $\Ubt(\xb,t)$, so that
\begin{equation}
\Ub = \Ubt(\xb,t) + w_0(r,t)\eb_r,
\end{equation}
We used $w_0$ to provide an estimate of the base state density evolution over a time step.

Our new temporal integration scheme does not use this splitting, and is much simpler than the scheme from Paper V.
In this new scheme we use a simpler, predictor-corrector approach to the base state density and pressure that no longer requires the complex algorithm from Paper V, yet still retains the same overall second-order accuracy.
Also, we now evolve the full-state velocity directly rather than the perturbational velocity, eliminating the need to compute additional forcing terms that arise due to the splitting.

>>>>

\subsection{Spatial Discretization}\label{Sec:Spatial}
\MarginPar{needs work}
MAESTRO contains variables defined on Cartesian grids 
one-dimensional base state stored on radial arrays.

@@

\subsection{Temporal Integration Scheme}\label{Sec:Temporal Integration Scheme}
Previously we adopted an approach where we split the velocity into a base state component, $w_0(r,t)$, 
and a local velocity $\Ubt(\xb,t)$, so that
\begin{equation}
\Ub = \Ubt(\xb,t) + w_0(r,t)\eb_r,
\end{equation}
We used $w_0$ to provide an estimate of the base state density evolution over a time step.
Our new temporal integration scheme does not use this splitting, and is much simpler than the scheme from Paper V since we evolve the velocity directly rather than the perturbational and base state velocity separately.  This eliminates
Additionally, the new scheme uses a simpler predictor-corrector approach to the base state density and pressure that no longer requires the complex algorithm from Paper V, yet still retains the same overall second-order accuracy.

### 2019-04-02 MAESTROeX_I/results.tex

In the left panel of Figure \ref{fig:scaling} we compare the wallclock time per time step as a function of total core count (in this case, the total number of OpenMP threads) for the original FBoxLib-based MAESTRO implementation to the AMReX MAESTROeX implementation.
These tests were performed using the original temporal integration strategy in \cite{MAESTRO_V}, noting that the new temporal integration scheme gives very similar performance results.\MarginPar{quantify this?... and also irregular dr case?}
We also include a plot of MAESTROeX without base state evolution.
Comparing the original and new implementations, we see similar scaling results except for the largest simulation, where MAESTROeX performs better.
We see that the increase in wallclock time from the smallest to largest simulation is roughly 42\%.
We also note that without base state evolution, the code runs 14\% faster for small problems, and scale mush better
However, for the case without base state evolution, the overall runtime is faster and the increase in wallclock time from the smallest to largest simulation is roughly 13\%.
This is quite remarkable since there are 3 linear solver per time step (2 cell-centered Poisson solves used in the MAC projection, and a nodal Poisson solve used to compute the updated cell-centered velocities).
Contrary to our prior assumptions, the linear solves are not the primary scaling bottleneck in this code.
In the right panel of Figure \ref{fig:scaling}, we isolate the wallclock time required for these linear solves and see that (i) the linear solves only use 20-23\% of the total computational time, and (ii) the increase in the solver wallclock time from the smallest to largest simulation is only 28\%.
Further profiling reveals that the primary scaling bottleneck is the average operator.
The averaging operator requires binning the sum of Cartesian data onto one-dimensional arrays holding every possible mapping radius.
This amounts to at least 24,384 points (for the $256^3$ simulation) up to 883,584 point (for the $1536^3$ simulation).
The averaging operator requires a global sum reduction over all processors, and the communication of this data is the primary scaling bottleneck.
For the simulation with base state evolution, this averaging operator is only called once per time step (as opposed to 14 times per time step when base state evolution is included).
The difference in total wallclock times with and without base state evolution is almost entirely due to the averaging.
Note that as expected, advection, reactions, and calls to the equation of state scale almost perfectly, since there is only a single parallel communication call to fill ghost cells.

>>>>

In the left panel of Figure \ref{fig:scaling} we compare the wallclock time per time step as a function of total core count (in this case, the total number of OpenMP threads) for the original FBoxLib-based MAESTRO implementation to the AMReX MAESTROeX implementation.
These tests were performed using the original temporal integration strategy in \cite{MAESTRO_V}, noting that the new temporal integration with and without the irregular base state gives essentially the same results.
We also include a plot of MAESTROeX without base state evolution.
Comparing the original and new implementations, we see similar scaling results except for the largest simulation, where MAESTROeX performs better.
We see that the increase in wallclock time from the smallest to largest simulation is roughly 42\%.
We also note that without base state evolution, the code runs 14\% faster for small problems, and scale mush better
However, for the case without base state evolution, the overall runtime is faster and the increase in wallclock time from the smallest to largest simulation is roughly 13\%.
This is quite remarkable since there are 3 linear solver per time step (2 cell-centered Poisson solves used in the MAC projection, and a nodal Poisson solve used to compute the updated cell-centered velocities).
Contrary to our prior assumptions, the linear solves are not the primary scaling bottleneck in this code.
In the right panel of Figure \ref{fig:scaling}, we isolate the wallclock time required for these linear solves and see that (i) the linear solves only use 20-23\% of the total computational time, and (ii) the increase in the solver wallclock time from the smallest to largest simulation is only 28\%.
Further profiling reveals that the primary scaling bottleneck is the average operator.
The averaging operator requires binning the sum of Cartesian data onto one-dimensional arrays holding every possible mapping radius.
This amounts to at least 24,384 double precision values (for the $256^3$ simulation) up to 883,584 values (for the $1536^3$ simulation).
The averaging operator requires a global sum reduction over all processors, and the communication of this data is the primary scaling bottleneck.
For the simulation with base state evolution, this averaging operator is only called once per time step (as opposed to 14 times per time step when base state evolution is included).
The difference in total wallclock times with and without base state evolution is almost entirely due to the averaging.
Note that as expected, advection, reactions, and calls to the equation of state scale almost perfectly, since there is only a single parallel communication call to fill ghost cells.
